Analytical Predictions of Explanet Obliquities Generated by Planet-Disk Interactions

### 2019-10-24 writeups/Su_disk_disp.tex
10/24 meeting

    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\langle#1\rangle}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator*{\med}{med}

@@

In our work, we provide an analytical framework for understanding the final
planetary obliquity for arbitrary initial spin-disk misalignment angles. Our
analytical work and independent numerical simulations are in agreement with
\citet{millholland_disk} but provide a more complete characterization of
possible behavior.

@@

Much of this treatment parallels that of \citet{anderson2018teeter} and
\citet{millholland_disk}. We consider an oblate star of mass $M_\star$ hosting a
planet of mass $M_p$, radius $R_p$, at semimajor axis $a_p$, and with spin
angular frequency $\Omega_p$, and a protoplanetary disk of mass $M_d$ at some
characteristic distance $r_d$ from the host star \citep[see][for a power-law
disk profile]{millholland_disk}. Denote $\hat{s} \equiv \vec{S} / S$ spin unit
vector of planet, $\hat{l} \equiv \vec{L} / L$ orbital angular momentum unit
vector of planet, and $\hat{l}_d \equiv \vec{L}_d / L_d$ angular momentum of the
surrounding disc. We approximate $S \ll L \ll L_d$, so $\hat{l}_d$ is
approximately constant and $\hat{l}$ experiences negligible backreaction torque
from $\hat{s}$. The quadrupole-order secular equations of motion for $\hat{s}$,
$\hat{l}$ and $\hat{l}_d$ then become very simple
\begin{align}
    \rd{\hat{s}}{t} &= \omega_{sl} \p{\hat{s} \cdot \hat{l}}
        \p{\hat{s} \times \hat{l}},\label{eq:dsdt}\\
    \rd{\hat{l}}{t} &= \omega_{ld}\p{\hat{l} \cdot \hat{l}_d}
        \p{\hat{l} \times \hat{l}_d}\label{eq:dldt}.
\end{align}
where
\begin{align}
    \omega_{sl} &\equiv \frac{3k_{qp}}{2k_p} \frac{M_\star}{m_p}
        \p{\frac{R_p}{a_p}}^3 \Omega_p,\label{eq:wsl}\\
    \omega_{ld} &\equiv \frac{3M_d}{4M_\star}\p{\frac{a_p}{r_d}}^3 n
        .\label{eq:wld}
\end{align}
% = 1/(152600 yr) (kp = 6kqp, Mstar = msun, mp = mJ, Rp = RJ, ap = AU, Wp = 2pi
% / (10 days))
% = 1 / (573362 yr) (Md/Mstar = 0.01, ap/rd = 1/30)
In \autoref{eq:wsl}, $k_{qp}$ and $k_p$ are related to the planet's quadrupole
moment and moment of inertia respectively \citep[see][]{lai2018}, and $n \equiv
\sqrt{GM_\star/a_p^3}$ is the planet's orbital mean motion. The three angles
formed by the three vectors are denoted
\begin{align}
    \theta &\equiv \arccos \hat{s} \cdot \hat{l},\\
    \theta_{sd} &\equiv \arccos \hat{s} \cdot \hat{l}_d,\\
    I &\equiv \arccos \hat{l} \cdot \hat{l}_d.
\end{align}

Let us transform to frame rotating about $\hat{l}_p$ with frequency $g \equiv
-\omega_{ld}\cos I < 0$, then $\hat{l}$ is fixed. In this frame, $\hat{l}_p$
also remains fixed and $\hat{s}$ evolves as:
\begin{equation}
    \rd{\hat{s}}{t} = \alpha \p{\hat{s} \cdot \hat{l}}
            \p{\hat{s} \times \hat{l}}
        + g\p{\hat{s} \times \hat{l}_d},
\end{equation}
where $\alpha \equiv \omega_{sl} > 0$ follows standard notation. By defining
$\eta$ and nondimensionalized time $\tau$ as
\begin{align}
    \eta \equiv{}& \frac{\abs{g}}{\alpha}\label{eq:eta},\\
        ={}& 3.75 \p{\frac{k_p}{6k_{qp}}}
            \p{\frac{m_p}{0.001 M_\odot}}
            \p{\frac{R_p}{R_J}}^{-3}
            \p{\frac{\Omega_p}{2\pi/\p{10\;\mathrm{days}}}}^{-1}\nonumber\\
        &\p{\frac{M_d}{0.01 M_{\odot}}}
            \p{\frac{a_p}{1\;\mathrm{AU}}}^{9/2}
            \p{\frac{r_d}{30 \;\mathrm{AU}}}^{-3}
            \p{\frac{M_\star}{1M_{\odot}}}^{-3/2}
            \cos I,\\
    \tau \equiv{}& \alpha t,
\end{align}
we can non-dimensionalize the equation of motion as
\begin{equation}
    \rd{\hat{s}}{\tau} = \p{\hat{s} \cdot \hat{l}}
            \p{\hat{s} \times \hat{l}}
        - \eta\p{\hat{s} \times \hat{l}_d}. \label{eq:dsdt_base}
\end{equation}

Following~\cite{millholland_disk}, we allow $g$ to vary in time owing to a
disk with decaying mass
\begin{equation}
    M_d(t) = M_d(0)e^{-t/t_d}.
\end{equation}
As $\alpha$ is held constant, this equates to $\eta$ decaying as $\rd{\eta}{t} =
-\frac{\eta}{t_d}$ as well. We nondimensionalize:
\begin{align}
    \epsilon \equiv{}& \frac{1}{\alpha t_d},\\
        ={}& 0.015 \p{\frac{k_p}{6k_{qp}}}
            \p{\frac{m_p/M_\star}{0.001}}
            \p{\frac{a_p}{1\;\mathrm{AU}}}^3\nonumber\\
        &\p{\frac{R_p}{R_J}}^{-3}
            \p{\frac{\Omega_p}{2\pi/\p{10\;\mathrm{days}}}}
            \p{\frac{t_d}{10\;\mathrm{Myr}}}^{-1},\\
    \rd{\eta}{\tau} ={}& -\epsilon \eta.\label{eq:deta_dt}
\end{align}
\autoref{eq:dsdt_base} and \autoref{eq:deta_dt} together constitute our system
of study.

@@

Spin states satisfying $\rd{\hat{s}}{t} = 0$ are referred to as \emph{Cassini
States} (CSs). There are either two or four solutions, depending on the values
of $\eta, I$. Following standard convention and nomenclature (see
\autoref{fig:cs_locs}), CSs 1, 3, 4 have $\phi = 0, \theta < 0$ and CS2 has
$\phi = \pi, \theta > 0$. Of these, CSs 1, 2, 3 are stable while CS4 is unstable
(see \autoref{s:local_dynamics}). When $\eta < \eta_c$, where
\begin{equation}
    \eta_c \equiv \p{\sin^{2/3}I + \cos^{2/3}I}^{-3/2},
\end{equation}
all four CSs exist, and when $\eta > \eta_c$, only CSs 2, 3 exist
\citep{henrard1987,ward2004I}. The CS obliquities as a function of $\eta$ are
shown in \autoref{fig:cs_locs}.
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/99_misc/2_cs_locs.png}
    \caption{Cassini state obliquities as a function of $\eta$. Note that
    $\theta \in [-\pi, \pi]$ is the traditional definition of the polar angle
    \citep[see e.g.][]{colombo1966,peale1969,henrard1987}.}\label{fig:cs_locs}
\end{figure}

@@

When $\eta < \eta_c$, CS4 exists and is a saddle point. The two trajectories
originating and ending at CS4 are the only two infinite-period orbits in phase
space. Together, these two critical trajectories are referred to as the
\emph{separatrix} and divide phase space into three zones. The separatrix, three
zones, and their relations to the CSs are shown via a contour plot of the
Hamiltonian of the system
\begin{equation}
    \mathcal{H} = -\frac{1}{2}\p{\hat{s} \cdot \hat{l}}^2
            + \eta \p{\hat{s} \cdot \hat{l}_d}.\label{eq:H}
\end{equation}
Such a contour plot is given in \autoref{fig:eq_1contours}. Note that
trajectories in zone $II$ librate about CS2 while those in zones $I$/$III$
circulate.
\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{../initial/0_eta/1contours_flip.png}
    \caption{Contour plot of $\mathcal{H}\p{\phi, \cos \theta}$ as given in
    \autoref{eq:H}, where warmer colors denote more positive values. The black
    solid line is the separatrix, which only exists for $\eta < \eta_c$. The
    three zones, divided by the separatrix, are labeled. The Cassini states are
    labeled and have the same colors as \autoref{fig:cs_locs}. The interior of
    the separatrix, shaded in grey, is formally only defined for $\eta <
    \eta_c$, but we may identify the phase space trajectories that flow into
    zone $II$ when evolved forward in time; this is the shaded region in the top
    left plot, bounded by the dotted black line.}\label{fig:eq_1contours}
\end{figure*}

The unsigned areas of the three zones are known exactly
\citep{henrard1987,ward2004I}. Defining
\begin{align*}
    z_0 &= \eta\cos I, &
    \chi &= \sqrt{-\frac{\tan^3\theta_4}{\tan I} - 1},\\
    \rho &= \chi \frac{\sin^2 \theta_4\cos \theta_4}{
        \chi^2 \cos^2\theta_4 + 1},&
    T &= 2\chi \frac{\cos \theta_4}{
        \chi^2 \cos^2\theta_4 - 1},\\
\end{align*}
the areas for $\eta < \eta_c$ are given by
\begin{subequations}\label{se:area_ward}
    \begin{align}
        A_{II} &= 8\rho + 4\arctan T - 8z_0 \arctan \frac{1}{\chi},\\
        A_I &= 2\pi\p{1 - z_0} - \frac{A_2}{2},\\
        A_{III} &= 2\pi\p{1 + z_0} - \frac{A_2}{2}.
    \end{align}
\end{subequations}
These are plotted as a function of $\eta$ in \autoref{fig:eq_areas}. Note that
the zones are not formally defined for $\eta > \eta_c$, but a natural extension
exists: evolve an initial point $p$ under adiabatic decrease of $\eta > \eta_c$
until the separatrix appears at $\eta = \eta_c$, then identify $p$ with the zone
it is in at $\eta_c$. Since phase space area is conserved under adiabatic
evolution, this implies $A_i\p{\eta > \eta_c} = A_i(\eta_c)$. This extension is
also reflected in the dashed boundary in \autoref{fig:eq_1contours}.
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/99_misc/1_areas.png}
    \caption{Plot of fractional areas of each of the zones $A_{i}(\eta) / 4\pi$
    as given by \autoref{se:area_ward}. Dotted lines correspond to small $\eta$
    approximations discussed in \autoref{s:ad_approx}. Dashed lines for $\eta >
    \eta_c$ effective values of $A_{II}, A_{III}$ for $\eta > \eta_c$, denoting
    the area of points that would flow into either area under adiabatic decrease
    of $\eta$ from $\eta > \eta_c$.}\label{fig:eq_areas}
\end{figure}

@@

One criterion for adiabaticity is that the separatrix crossing timescale be
longer than the the resonant libration period \citep{ward2004II}. The resonant
libration period for a nearly separatrix-crossing orbit can be approximated by
the libration period near CS2, resulting in constraint
\begin{align}
    -\at{\frac{1}{\eta}\rd{\eta}{\tau}}_{cross} = \epsilon &\lesssim
            \frac{1}{T_{lib}},\\
        &\lesssim \frac{1}{2\pi}\sqrt{\eta_\star\sin I \sin \theta_2
            \p{1 + \eta_\star \sin I \csc^3 \theta_2}},
            \label{eq:ad_constr}
\end{align}
where $\theta_2$ is the obliquity at CS2. This formula differs from that
presented in \citet{millholland_disk}, where the $\csc^3\theta_2$ term is
neglected and a square root is missing. See \autoref{s:local_dynamics} for a
detailed derivation of $T_{lib}$ and comparison to existing work.

@@

We consider the evolution of a system with arbitrary initial $\theta_{sd, i}
\equiv \at{\theta_{sd}}_{t = 0}$ and $\eta_i \equiv \at{\eta}_{t = 0} \gg 1$. We
are interested in the distribution of final obliquities after $\eta$ decreases
to $\eta_f \ll 1$ (after the disk has dissipated to negligible mass). While it
is likely that $\theta_{sd, i} \approx 0$ as the planet forms in the disk, we
explore all $\theta_{sd, i} \in [0, \pi]$ for completeness.

We evenly sample $101$ values of $\theta_{sd, i}$, and for each $\theta_{sd, i}$
value, we pick $101$ $\hat{s}$ initial conditions approximately on the ring
fixed by $\theta_{sd, i}$\footnote{For each $\theta_{sd, i}$ we choose an
initial condition on the libration cycle going through $\p{\theta_2 +
\theta_{sd, i}, \phi_2}$ (where $\p{\theta_2, \phi_2}$ are the coordinates of
CS2), such that all initial conditions for a given $\theta_{sd, i}$ enclose the
same initial phase space; see \autoref{sss:a_evo}.}. To be concrete, we choose
$\eta_i = 10\eta_c$. For each initial condition, we evolve
Equations~\ref{eq:dsdt_base} and~\ref{eq:deta_dt} until $\eta = 10^{-5}$. For
such small $\eta$, $\hat{s}$ precesses about $\hat{l}$ with uniform final
obliquity $\theta_f$. This transfer function $\theta_f\p{\theta_{sd, i}}$ is our
primary result, and is shown for $I = 5^\circ, 10^\circ, 20^\circ$ in
\autoref{fig:ad_ensemble}, \autoref{fig:3_ensemble_10_35}, and
\autoref{fig:3_ensemble_20_35} respectively. The blue dots represent the results
of the simulation, which are in excellent agreement with the colored tracks.
These colored tracks are calculated semi-analytically using methods explained in
the following section.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/2_toy2/3_ensemble_05_35.png}
    \caption{Top: $\theta_{f}\p{\theta_{sd, i}}$, overlaid with semi-analytic
    predictions of the $\theta_{ f}$ for each of the four nontrivial
    dynamical tracks in colored lines. Black dotted lines correspond to
    small-$\theta_{sd, i}$ predictions derived in \autoref{s:ad_approx}.
    Bottom: Semi-analytic probabilities of each of the dynamical tracks for each
    $\theta_{sd, i}$. The five regimes of $\theta_{sd, i}$ values demarcated by
    the vertical dashed black lines correspond to the five regimes of $A_i$
    calculated in \autoref{ss:zone_transitions}. Black dotted lines correspond to
    simple analytic estimates given in
    \autoref{s:ad_approx}.}\label{fig:ad_ensemble}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/2_toy2/3_ensemble_10_35.png}
    \caption{Same as \autoref{fig:ad_ensemble} but for $I =
    10^\circ$ and with fewer annotations.}\label{fig:3_ensemble_10_35}
\end{figure}
\begin{figure}
    \includegraphics[width=0.5\textwidth]{../initial/2_toy2/3_ensemble_20_35.png}
    \caption{Same as \autoref{fig:3_ensemble_10_35} but for $I = 20^\circ$. The
    $III \to II \to I$ track is very short since the maximum $A_{II}$ is very
    near $\eta_c$ for $I = 20^\circ$.}\label{fig:3_ensemble_20_35}
\end{figure}

@@

The evolutionary tracks that govern the behavior of $\theta_f\p{\theta_{sd,
i}}$ correspond to particular sequences of separatrix crossings. Coarsely, they
can be understood using a combination of (i) how the enclosed phase space area
by the trajectory evolves across each separatrix crossing, and (ii) the
associated probabilities with each separatrix crossing.

\subsubsection{Governing Principles: Evolution of $A$}\label{sss:a_evo}

>>>>

    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\langle#1\rangle}
    \newcommand*{\bm}[1]{\boldsymbol{\mathbf{#1}}}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator*{\med}{med}

@@

In our work, we provide an analytical framework for understanding the final
planetary obliquity for arbitrary initial spin-disk misalignment angles. Our
analytical work and independent numerical simulations are in agreement with
\citet{millholland_disk} but provide a more complete characterization of
possible behavior. While we solve a specific dynamics problem, our analysis
is general and has a wide range of applicability to general resonance
encounters.

@@

Much of this treatment parallels that of \citet{anderson2018teeter} and
\citet{millholland_disk}. We consider an oblate star of mass $M_\star$ hosting a
planet of mass $M_{\rm p}$, radius $R_{\rm p}$, at semimajor axis $a_{\rm p}$,
and with spin angular frequency $\Omega_{\rm p}$, and a protoplanetary disk of
mass $M_{\rm d}$ at some characteristic distance $r_{\rm d}$ from the host star
\citep[see][for a power-law disk profile]{millholland_disk}. Denote $\hat{s}
\equiv \bm{S} / S$ spin unit vector of planet, $\hat{l} \equiv \bm{L} / L$
orbital angular momentum unit vector of planet, and $\hat{l}_{\rm d} \equiv
\bm{L}_{\rm d} / L_{\rm d}$ angular momentum of the surrounding disc. We
approximate $S \ll L \ll L_{\rm d}$, so $\hat{l}_{\rm d}$ is approximately
constant and $\hat{l}$ experiences negligible backreaction torque from
$\hat{s}$. The quadrupole-order secular equations of motion for $\hat{s}$,
$\hat{l}$ and $\hat{l}_{\rm d}$ then become very simple
\begin{align}
    \rd{\hat{s}}{t} &= \omega_{\rm sl} \p{\hat{s} \cdot \hat{l}}
        \p{\hat{s} \times \hat{l}},\label{eq:dsdt}\\
    \rd{\hat{l}}{t} &= \omega_{\rm ld}\p{\hat{l} \cdot \hat{l}_{\rm d}}
        \p{\hat{l} \times \hat{l}_{\rm d}}\label{eq:dldt}.
\end{align}
where
\begin{align}
    \omega_{\rm sl} &\equiv \frac{3k_{\rm qp}}{2k_{\rm p}} \frac{M_\star}{m_{\rm
        p}} \p{\frac{R_{\rm p}}{a_{\rm p}}}^3 \Omega_{\rm p},\label{eq:wsl}\\
    \omega_{\rm ld} &\equiv \frac{3M_{\rm d}}{4M_\star}\p{\frac{a_{\rm
        p}}{r_{\rm d}}}^3 n .\label{eq:wld}
\end{align}
% = 1/(152600 yr) (kp = 6kqp, Mstar = msun, mp = mJ, Rp = RJ, ap = AU, Wp = 2pi
% / (10 days))
% = 1 / (573362 yr) (Md/Mstar = 0.01, ap/rd = 1/30)
In \autoref{eq:wsl}, $k_{\rm qp}$ and $k_{\rm }$ are related to the planet's
quadrupole moment and moment of inertia respectively \citep[see][]{lai2018}, and
$n \equiv \sqrt{GM_\star/a_{\rm p}^3}$ is the planet's orbital mean motion. The
three angles formed by the three vectors are denoted
\begin{align}
    \theta &\equiv \arccos \hat{s} \cdot \hat{l},\\
    \theta_{\rm sd} &\equiv \arccos \hat{s} \cdot \hat{l}_{\rm d},\\
    I &\equiv \arccos \hat{l} \cdot \hat{l}_{\rm d}.
\end{align}

Let us transform to frame rotating about $\hat{l}_{\rm p}$ with frequency $g
\equiv -\omega_{\rm ld}\cos I < 0$, then $\hat{l}$ is fixed. In this frame,
$\hat{l}_{\rm p}$ also remains fixed and $\hat{s}$ evolves as:
\begin{equation}
    \rd{\hat{s}}{t} = \alpha \p{\hat{s} \cdot \hat{l}}
            \p{\hat{s} \times \hat{l}}
        + g\p{\hat{s} \times \hat{l}_{\rm d}},
\end{equation}
where $\alpha \equiv \omega_{\rm sl} > 0$ follows standard notation. By defining
$\eta$ and nondimensionalized time $\tau$ as
\begin{align}
    \eta \equiv{}& \frac{\abs{g}}{\alpha}\label{eq:eta},\\
        ={}& 3.75 \p{\frac{k_{\rm p}}{6k_{\rm qp}}}
            \p{\frac{m_{\rm p}}{0.001 M_\odot}}
            \p{\frac{R_{\rm p}}{R_J}}^{-3}
        &\p{\frac{M_{\rm d}}{0.01 M_{\odot}}}
            \p{\frac{a_{\rm p}}{1\;\mathrm{AU}}}^{9/2}
            \p{\frac{r_{\rm d}}{30 \;\mathrm{AU}}}^{-3}
            \p{\frac{M_\star}{1M_{\odot}}}^{-3/2}
            \cos I,\\
    \tau \equiv{}& \alpha t,
\end{align}
we can non-dimensionalize the equation of motion as
\begin{equation}
    \rd{\hat{s}}{\tau} = \p{\hat{s} \cdot \hat{l}}
            \p{\hat{s} \times \hat{l}}
        - \eta\p{\hat{s} \times \hat{l}_{\rm d}}. \label{eq:dsdt_base}
\end{equation}

Following~\cite{millholland_disk}, we allow $g$ to vary in time owing to a
disk with decaying mass
\begin{equation}
    M_{\rm d}(t) = M_{\rm d}(0)e^{-t/t_{\rm d}}.
\end{equation}
As $\alpha$ is held constant, this equates to $\eta$ decaying as $\rd{\eta}{t} =
-\frac{\eta}{t_{\rm d}}$ as well. We nondimensionalize:
\begin{align}
    \epsilon \equiv{}& \frac{1}{\alpha t_{\rm d}},\\
        ={}& 0.015 \p{\frac{k_{\rm p}}{6k_{\rm qp}}}
            \p{\frac{m_{\rm p}/M_\star}{0.001}}
            \p{\frac{a_{\rm p}}{1\;\mathrm{AU}}}^3\nonumber\\
        &\p{\frac{R_{\rm p}}{R_J}}^{-3}
            \p{\frac{\Omega_{\rm p}}{2\pi/\p{10\;\mathrm{days}}}}
            \p{\frac{t_{\rm d}}{10\;\mathrm{Myr}}}^{-1},\\
    \rd{\eta}{\tau} ={}& -\epsilon \eta.\label{eq:deta_dt}
\end{align}
\autoref{eq:dsdt_base} and \autoref{eq:deta_dt} together constitute our system
of study.

In the next two subsections, we further discuss theoretical background relevant
to our analysis of the evolution of this system. They are not essential to
reproducing the results presented in \autoref{ss:ad_ensemble}.

@@

Spin states satisfying $\rd{\hat{s}}{t} = 0$ are referred to as \emph{Cassini
States} (CSs). There are either two or four solutions, depending on the values
of $\eta, I$. Following standard convention and nomenclature (see
\autoref{fig:cs_locs}), CSs 1, 3, 4 have $\phi = 0, \theta < 0$ and CS2 has
$\phi = \pi, \theta > 0$. Of these, CSs 1, 2, 3 are stable while CS4 is unstable
(see \autoref{s:local_dynamics}). When $\eta < \eta_{\rm c}$, where
\begin{equation}
    \eta_{\rm c} \equiv \p{\sin^{2/3}I + \cos^{2/3}I}^{-3/2},
\end{equation}
all four CSs exist, and when $\eta > \eta_{\rm c}$, only CSs 2, 3 exist
\citep{henrard1987,ward2004I}. The CS obliquities as a function of $\eta$ are
shown in \autoref{fig:cs_locs}.
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/99_misc/2_cs_locs.png}
    \caption{Cassini state obliquities as a function of $\eta$. Note that
    $\theta \in [-\pi, \pi]$ is the traditional definition of the polar angle
    \citep[see e.g.][]{colombo1966,peale1969,henrard1987}. The vertical dashed
    line indicates $\eta_{\rm c}$, where CS1 and CS4 merge and annihilate, and the
    horizontal dashed lines indicate $\theta = I, 180^\circ - I$ the asymptotic
    behavior for CS2, 3 for $\eta \gg \eta_{\rm c}$.}\label{fig:cs_locs}
\end{figure}

@@

When $\eta < \eta_{\rm c}$, CS4 exists and is a saddle point. The two trajectories
originating and ending at CS4 are the only two infinite-period orbits in phase
space. Together, these two critical trajectories are referred to as the
\emph{separatrix} and divide phase space into three zones. The separatrix, three
zones, and their relations to the CSs are shown via a contour plot of the
Hamiltonian of the system
\begin{equation}
    \mathcal{H} = -\frac{1}{2}\p{\hat{s} \cdot \hat{l}}^2
            + \eta \p{\hat{s} \cdot \hat{l}_{\rm d}}.\label{eq:H}
\end{equation}
Such a contour plot is given in \autoref{fig:eq_1contours}. Note that
trajectories in zone $II$ librate about CS2 while those in zones $I$/$III$
circulate.
\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{../initial/0_eta/1contours_flip.png}
    \caption{Contour plot of $\mathcal{H}\p{\phi, \cos \theta}$ as given in
    \autoref{eq:H}, where warmer colors denote more positive values. The black
    solid line is the separatrix, which only exists for $\eta < \eta_{\rm c}$. The
    three zones, divided by the separatrix, are labeled. The Cassini states are
    labeled and have the same colors as \autoref{fig:cs_locs}. The interior of
    the separatrix, shaded in grey, is formally only defined for $\eta <
    \eta_{\rm c}$, but we may identify the phase space trajectories that flow into
    zone $II$ when evolved forward in time; this is the shaded region in the top
    left plot, bounded by the dotted black line.}\label{fig:eq_1contours}
\end{figure*}

The unsigned areas of the three zones are known exactly
\citep{henrard1987,ward2004I}. Defining
\begin{subequations}\label{se:area_ward}
    \begin{align}
        z_0 &= \eta\cos I, &
        \chi &= \sqrt{-\frac{\tan^3\theta_4}{\tan I} - 1},\\
        \rho &= \chi \frac{\sin^2 \theta_4\cos \theta_4}{
            \chi^2 \cos^2\theta_4 + 1},&
        T &= 2\chi \frac{\cos \theta_4}{
            \chi^2 \cos^2\theta_4 - 1},\\
    \end{align}
    the areas for $\eta < \eta_{\rm c}$ are given by
    \begin{align}
        A_{\rm II} &= 8\rho + 4\arctan T - 8z_0 \arctan \frac{1}{\chi},\\
        A_{\rm I} &= 2\pi\p{1 - z_0} - \frac{A_2}{2},\\
        A_{\rm III} &= 2\pi\p{1 + z_0} - \frac{A_2}{2}.
    \end{align}
\end{subequations}
These are plotted as a function of $\eta$ in \autoref{fig:eq_areas}. Note that
the zones are not formally defined for $\eta > \eta_{\rm c}$, but a natural
extension exists: evolve an initial point $p$ under adiabatic decrease of $\eta
> \eta_{\rm c}$ until the separatrix appears at $\eta = \eta_{\rm c}$, then
identify $p$ with the zone it is in at $\eta_{\rm c}$. Since phase space area is
conserved under adiabatic evolution, this implies $A_i\p{\eta > \eta_{\rm c}} =
A_i(\eta_{\rm c})$. This extension is also reflected in the dashed boundary in
\autoref{fig:eq_1contours}.
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/99_misc/1_areas.png}
    \caption{Plot of fractional areas of each of the zones $A_{\rm i}(\eta) /
    4\pi$ as given by \autoref{se:area_ward}. Dotted lines correspond to small
    $\eta$ approximations discussed in \autoref{s:ad_approx}. Dashed lines for
    $\eta > \eta_{\rm c}$ effective values of $A_{\rm II}, A_{\rm III}$ for
    $\eta > \eta_{\rm c}$, denoting the area of points that would flow into
    either area under adiabatic decrease of $\eta$ from $\eta > \eta_{\rm
    c}$.}\label{fig:eq_areas}
\end{figure}

@@

One criterion for adiabaticity is that the separatrix crossing timescale be
longer than the the resonant libration period \citep{ward2004II}. The resonant
libration period for a nearly separatrix-crossing orbit is approximated by
the libration period near CS2 in \citet{millholland_disk}. We find that this
criterion performs remarkably well despire the expectation that the separatrix
crossing orbit should be much longer than libration near CS2 (approaching
infinity as $\epsilon \to 0$). One justification for this is that the libration
period about CS2 is the longest timescale in the system (see
\autoref{fig:lambda2})

This constraint can be computed
\begin{align}
    -\at{\frac{1}{\eta}\rd{\eta}{\tau}}_{\rm cross} = \epsilon &\lesssim
            \frac{1}{T_{\rm lib}},\\
        &\lesssim \frac{1}{2\pi}\sqrt{\eta_\star\sin I \sin \theta_2
            \p{1 + \eta_\star \sin I \csc^3 \theta_2}},
            \label{eq:ad_constr}
\end{align}
where $\theta_2$ is the obliquity at CS2. This formula differs from that
presented in \citet{millholland_disk}, where the $\csc^3\theta_2$ term is
neglected and a square root is missing.

@@

We consider the evolution of a system with arbitrary initial $\theta_{\rm sd, i}
\equiv \at{\theta_{\rm sd}}_{t = 0}$ and $\eta_i \equiv \at{\eta}_{t = 0} \gg
1$. We are interested in the distribution of final obliquities after $\eta$
decreases to $\eta_f \ll 1$ (after the disk has dissipated to negligible mass).
While it is likely that $\theta_{\rm sd, i} \approx 0$ as the planet forms in
the disk, we explore all $\theta_{\rm sd, i} \in [0, \pi]$ for completeness.

We evenly sample $101$ values of $\theta_{\rm sd, i}$, and for each $\theta_{\rm
sd, i}$ value, we pick $101$ $\hat{s}$ initial conditions approximately on the
ring fixed by $\theta_{\rm sd, i}$\footnote{For each $\theta_{\rm sd, i}$ we
choose an initial condition on the libration cycle going through $\p{\theta_2 +
\theta_{\rm sd, i}, \phi_2}$ (where $\p{\theta_2, \phi_2}$ are the coordinates
of CS2), such that all initial conditions for a given $\theta_{\rm sd, i}$
enclose the same initial phase space; see \autoref{sss:a_evo}.}. To be concrete,
we choose $\eta_i = 10\eta_{\rm c}$. For each initial condition, we evolve
Equations~\ref{eq:dsdt_base} and~\ref{eq:deta_dt} until $\eta = 10^{-5}$. For
such small $\eta$, $\hat{s}$ precesses about $\hat{l}$ with uniform final
obliquity $\theta_{\rm f}$. This transfer function $\theta_{\rm f}\p{\theta_{\rm
sd, i}}$ is our primary result, and is shown for $I = 5^\circ, 10^\circ,
20^\circ$ in \autoref{fig:ad_ensemble}, \autoref{fig:3_ensemble_10_35}, and
\autoref{fig:3_ensemble_20_35} respectively. The blue dots represent the results
of the simulation, which are in excellent agreement with the colored tracks.
These colored tracks are calculated semi-analytically using methods explained in
the following section.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/2_toy2/3_ensemble_05_35.png}
    \caption{Top: $\theta_{\rm f}\p{\theta_{\rm sd, i}}$, overlaid with
    semi-analytic predictions of the $\theta_{\rm  f}$ for each of the four
    nontrivial dynamical tracks in colored lines. Black dotted lines correspond
    to small-$\theta_{\rm sd, i}$ predictions derived in \autoref{s:ad_approx}.
    Bottom: Semi-analytic probabilities of each of the dynamical tracks for each
    $\theta_{\rm sd, i}$. The five regimes of $\theta_{\rm sd, i}$ values
    demarcated by the vertical dashed black lines correspond to the five regimes
    of $A_i$ calculated in \autoref{ss:zone_transitions}. Black dotted lines
    correspond to simple analytic estimates given in
    \autoref{s:ad_approx}.}\label{fig:ad_ensemble}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/2_toy2/3_ensemble_10_35.png}
    \caption{Same as \autoref{fig:ad_ensemble} but for $I =
    10^\circ$ and with fewer annotations.}\label{fig:3_ensemble_10_35}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/2_toy2/3_ensemble_20_35.png}
    \caption{Same as \autoref{fig:3_ensemble_10_35} but for $I = 20^\circ$. The
    $III \to II \to I$ track is very short since the maximum $A_{\rm II}$ is very
    near $\eta_{\rm c}$ for $I = 20^\circ$.}\label{fig:3_ensemble_20_35}
\end{figure}

@@

The evolutionary tracks that govern the behavior of $\theta_{\rm
f}\p{\theta_{\rm sd, i}}$ correspond to particular sequences of separatrix
crossings. Coarsely, they can be understood using a combination of (i) how the
enclosed phase space area by the trajectory evolves across each separatrix
crossing, and (ii) the associated probabilities with each separatrix crossing.

\subsubsection{Governing Principles: Evolution of Enclosed Phase Space
Area}\label{sss:a_evo}

First, we consider how the enclosed phase space area by a trajectory evolves
over time. In the absence of separatrix encounters, the enclosed phase space
area $\oint \cos\theta \;\mathrm{d}\phi$ is an adiabatic invariant. In
particular, we adopt convention where
\begin{equation}
    A \equiv \oint \p{1 - \cos \theta}\;\mathrm{d}\phi.\label{eq:a_oint}
\end{equation}
This definition of $A$ has the advantage of (i) being continuous at $\cos \theta
= 1$, and (ii) being easily expressible as combinations of the $A_i$ of
\autoref{se:area_ward}. The bounds of the integral are either a libration
($\phi$ returns to its original value with the same $\dot{q\phi}$ sign) or
circulation cycle ($\phi$ advances to $\phi \pm 2\pi$). Note lastly that $A$ has
simple physical interpretation of are enclosed on the unit sphere by $\hat{s}$
over one cycle measured relative to $+\hat{z}$:
\begin{equation}
    \int \int\limits_{\cos \theta(\phi)}^1
        \mathrm{d}\cos\theta\;\mathrm{d}\phi = \int \p{1 - \cos \theta(\phi)}
            \;\mathrm{d}\phi.
\end{equation}
